
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(readr)
require(tidyverse)
require(magrittr)
require(ggplot2)
require(gtsummary)
require(reshape2)
require(officer)
require(deSolve)
require(DiagrammeR)
```

##MODELLING AN EPIDEMIC
#Why do we model epidemics?
#Predict the current conditions
#Predict future progress of any given epidemic
#Understand the mechanisms behind the epidemic
#Understand which interventions might be successful

##How do we model epidemics?
#Identify key biological processes
#Abstract them into a mathematical model
#Investigate effect(s) of changing parameters

## SI MODEL
#Assume
#fixed population of constant size (S+I=N)
#All members of the population are in contact

#Rate at which a susceptible person gets infected depends upon:
#The number or proportion of infected persons
#The transmission rate β

##Rate of infection
#We are going to focus on frequency dependent transmission
#This is the most appropriate for models in humans
#Livestock models may use density dependent transmission


##Understanding the model
```{r}
grViz("digraph flowchart {
      graph [layout = dot,
       rankdir = LR]
      node [fontname = Helvetica, shape = oval,]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      # edge definitions with the node IDs
      tab1 -> tab2;
      
      }

      [1]: 'INFECTED'
      [2]: 'RECOVERED'
      ")
```



##1.Evidence synthesis
#As a pre-requisite to model building, we need to undertake evidence 
#synthesis exercise to extract required parameters from the literature or by conducting a study.

```{r}
initial_number_infected <- 1000
initial_number_recovered <- 0 
Duration_of_infectiousness <- 7 
recovery_rate <- 1/Duration_of_infectiousness
follow_up_duration <- 30

```


##2. Create model input using parameters
#Once parameters are obtained, they are placed into three major 
#parameter inputs viz initial values for each compartment, 
#transition parameters from one compartment to another, and the time 
#stamps for which modelling is required.

```{r}
initial_state_values <- c(I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(gamma = recovery_rate) 
times <- seq(from = 0, to = follow_up_duration, by = 1) 
```

##3. Create a model function using input values
#The model is created from the above mentioned parameters by specifying differential equations. In R, a function is an object so that the researcher is able to pass control to the function, along with arguments that may be necessary for the function to accomplish the actions. For performing differntial equations, we create a function with compartment as well as transition parameter at varied timestamps.
```{r}
cohort_model <- function(time, state, parameters) {     
  with(as.list(c(state, parameters)), {  
    dI <- -gamma * I
    dR <- gamma * I
    return(list(c(dI, dR)))                             
  })
}
```

##4. Return a dataframe as output using model equations
#A data frame is constructed by application of the differential model. Herein, Odinary Differential Equations (ode) function takes initial compartment values, timestamps, parameters and the model function as arguments.
```{r}
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))

```

##5. Exploratory Data Analysis of the output.

#To understand the output let us first see the dimensions of our dataframe
```{r}
dim(output)

#To look at the first 5 values of the data

head(output,5)
```
#Since, more than one column is representing the state of 
#the person (Infected or recovered), we need to transform 
#the data into tidy format. A tidy data format is said to 
#present when each column represents one variable, each row 
#has one observation and each cell has one value.

```{r}
dat <- output %>% pivot_longer(cols = c(I,R),
                               names_to = "state",
                               values_to = "value")
head(dat,10)  
```


#Visualization is a recommended exploratory data analysis method. 
#Let us see how infected and recovered compartments were changing in the model.

```{r}
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Infected and Recovered individuals")
```


##Increasing complexity
#To demonstrate increasing complexity, let us now develop a model 
#for a disease wherein three compartments viz susceptible, infected, 
#and recovered are present. There are no births (inflows) or deaths (outflows)
#in the model. Further, all those who are infected are also infectious and once recovered, 
#immunity is long lasting. The basic assumptions of compartment models such as homogenity 
#among others are also applicable.

##Visualization of the model
```{r}
grViz("digraph flowchart {
      graph [layout = dot,
       rankdir = LR]
      node [fontname = Helvetica, shape = oval,]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3'] 
      # edge definitions with the node IDs
      tab1 -> tab2;
      tab2 -> tab3;
      }

      [1]: 'Susceptible'
      [2]: 'Infected'
      [3]: 'Recovered'
      ")
```

##Adding changes in the IR model to develop SIR model
```{r}
## 1. Evidence synthesis
initial_number_susceptible <- 10000000 # Adding susceptibles to IR model
initial_number_infected <- 1 # Considering one infected at start time
initial_number_recovered <- 0 
Duration_of_infectiousness <- 7 
recovery_rate <- 1/Duration_of_infectiousness
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
```

```{r}
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 
```


```{r}
## 3. Create a model using input values
cohort_model <- function(time, state, parameters) {     
  with(as.list(c(state, parameters)), { 
    N <- S+I+R # Total population
    lambda <- beta * I/N #Calculating lambda as function of beta
    dS <- -lambda * S               
    dI <- lambda * S - gamma * I    
    dR <- gamma * I   
    return(list(c(dS, dI, dR)))                             
  })
}
```

```{r}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
```


```{r}
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                               names_to = "state",
                               values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered")

```



##Adding changes in the IR model to develop SIR model
```{r}
## 1. Evidence synthesis
initial_number_susceptible <- 10000000 # Adding susceptibles to IR model
initial_number_infected <- 1 # Considering one infected at start time
initial_number_recovered <- 0 
Duration_of_infectiousness <- 7 
recovery_rate <- 1/Duration_of_infectiousness
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
```

```{r}
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1)
```


```{r}
## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
  with(as.list(c(state, parameters)), { 
    N <- S+I+R # Total population
    lambda <- beta * I/N #Calculating lambda as function of beta
    dS <- -lambda * S               
    dI <- lambda * S - gamma * I    
    dR <- gamma * I   
    return(list(c(dS, dI, dR)))                             
  })
}
```

```{r}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
```

```{r}
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                               names_to = "state",
                               values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered")
```


##Increasing complexities: Addition of interventions.
#Let us now assume that we have a vaccine with 100% effectiveness against aquiring infection. Will the epidemic be averted by vaccinating 10% population? What other changes do you expect in the epidemic?
```{r}
## 1. Evidence synthesis
prop_vaccinated <- 0.1
initial_number_susceptible <- (1-prop_vaccinated)*10000000 
initial_number_infected <- 1 
initial_number_recovered <- prop_vaccinated*10000000
Duration_of_infectiousness <- 7 
recovery_rate <- 1/7
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
```


```{r}
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, 
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate,  
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 
```


```{r}
## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
   with(as.list(c(state, parameters)), { 
     N <- S+I+R # Total population
     lambda <- beta * I/N #Calculating lambda as function of beta
     dS <- -lambda * S               
     dI <- lambda * S - gamma * I    
     dR <- gamma * I   
   return(list(c(dS, dI, dR)))                             
   })
}
```


```{r}
## 4. Return a data frame as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
```

```{r}
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                        names_to = "state",
                        values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered at 10% vaccine coverage")

```

##Will the epidemic be averted by increasing vaccination coverage? What is the critical vaccination threshold?
#Critical Vaccination Threshold is defined as minimum percentage of population which needs to be vaccinated to prevent occurence of epidemic. Going manually by changing the vaccination coverage is a start point in understanding the concept.

##Modelling for 20% coverage:
```{r}
## 1. Evidence synthesis
prop_vaccinated <- 0.2
initial_number_susceptible <- (1-prop_vaccinated)*10000000 
initial_number_infected <- 1 
initial_number_recovered <- prop_vaccinated*10000000
Duration_of_infectiousness <- 7 
recovery_rate <- 1/7
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 

## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
   with(as.list(c(state, parameters)), { 
     N <- S+I+R # Total population
     lambda <- beta * I/N #Calculating lambda as function of beta
     dS <- -lambda * S               
     dI <- lambda * S - gamma * I    
     dR <- gamma * I   
   return(list(c(dS, dI, dR)))                             
   })
}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                        names_to = "state",
                        values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered at 20% vaccine coverage")

```


##Modelling for 30% coverage:
```{r}
## 1. Evidence synthesis
prop_vaccinated <- 0.3
initial_number_susceptible <- (1-prop_vaccinated)*10000000 
initial_number_infected <- 1 
initial_number_recovered <- prop_vaccinated*10000000
Duration_of_infectiousness <- 7 
recovery_rate <- 1/7
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 

## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
   with(as.list(c(state, parameters)), { 
     N <- S+I+R # Total population
     lambda <- beta * I/N #Calculating lambda as function of beta
     dS <- -lambda * S               
     dI <- lambda * S - gamma * I    
     dR <- gamma * I   
   return(list(c(dS, dI, dR)))                             
   })
}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                        names_to = "state",
                        values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered at 30% vaccine coverage")

```

##Modelling for 40% coverage:
```{r}
## 1. Evidence synthesis
prop_vaccinated <- 0.4
initial_number_susceptible <- (1-prop_vaccinated)*10000000 
initial_number_infected <- 1 
initial_number_recovered <- prop_vaccinated*10000000
Duration_of_infectiousness <- 7 
recovery_rate <- 1/7
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 

## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
   with(as.list(c(state, parameters)), { 
     N <- S+I+R # Total population
     lambda <- beta * I/N #Calculating lambda as function of beta
     dS <- -lambda * S               
     dI <- lambda * S - gamma * I    
     dR <- gamma * I   
   return(list(c(dS, dI, dR)))                             
   })
}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                        names_to = "state",
                        values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered at 40% vaccine coverage")

```

##Modelling for 50% coverage:
```{r}
## 1. Evidence synthesis
prop_vaccinated <- 0.5
initial_number_susceptible <- (1-prop_vaccinated)*10000000 
initial_number_infected <- 1 
initial_number_recovered <- prop_vaccinated*10000000
Duration_of_infectiousness <- 7 
recovery_rate <- 1/7
Reproduction_number <- 2.28 #R0== beta/gamma,
infection_rate <- Reproduction_number*(1/Duration_of_infectiousness) #beta
follow_up_duration <- 300
## 2. Create model input using parameters
initial_state_values <- c(S = initial_number_susceptible, # Adding susceptibles
                          I = initial_number_infected, 
                          R = initial_number_recovered) 

parameters <- c(beta = infection_rate, # Adding beta parameter 
                gamma = recovery_rate) 

times <- seq(from = 0, to = follow_up_duration, by = 1) 

## 3. Create a model using input values

cohort_model <- function(time, state, parameters) {     
   with(as.list(c(state, parameters)), { 
     N <- S+I+R # Total population
     lambda <- beta * I/N #Calculating lambda as function of beta
     dS <- -lambda * S               
     dI <- lambda * S - gamma * I    
     dR <- gamma * I   
   return(list(c(dS, dI, dR)))                             
   })
}
## 4. Return a dataframe as output using model equations
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = cohort_model,
                            parms = parameters))
## 5. Exploratory Data Analysis (Visualisation) of the output
dat <- output %>% pivot_longer(cols = c(S,I,R),
                        names_to = "state",
                        values_to = "value")
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "Changes in Suceptible, Infected and Recovered at 50% vaccine coverage")

```